---
title: "prepareBQEdata"
author: "HT"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(openxlsx)
library(tidyverse)
library(dplyr)
library(tidyr)
```

## Data description

This script prepares the previously extracted biological classifications (BQE from WISE-4 & WISE-2) to allow joining it with the water quality data (WQ from WISE-6) for selected supporting elements (SE), for both TRaC water categories (TW & CW).

* BQE - From WISE2, EQR & EQS aggregated by Site (yearly), except for EE, IE and SI for which only already aggregated data by WB was available at source, so the EQR & EQSaggWB data was also extracted (for all countries and for those 4 countries). Data was extracted for all years available.

* BQE - From WISE4, only WB aggregated EQS data (i.e., categories, not EQR) was available for the 1st and 2nd reporting cycle years 2010 and 2016.

## B. Import Biological QE data
### B1. Import BQE data for TraC Waters per year
```{r Import WISE4 Biota EQS classifications to Wide format}
#data split at Water Body level considering the years ]2009, 2016] leading to the 2nd reporting cycle 2016

#TW
dat_WFD_TW <- readRDS(here("Data","dat_W4_TW_BQE.rds")) #import biology

#check available reporting years
unique(dat_WFD_TW$cYear) # 2010, 2016

#all years
dat_WFD_TW.wide <- dat_WFD_TW %>% 
  select(euSurfaceWaterBodyCode,countryCode,qeCode,qeStatusOrPotentialValue,cYear,naturalAWBHMWB,surfaceWaterBodyTypeCode,
         surfaceWaterBodyIntercalibrationType,surfaceWaterBodyIntercalibrationTypeCode,cArea,cLength) %>% 
  pivot_wider(names_from = qeCode, values_from = qeStatusOrPotentialValue)

colnames(dat_WFD_TW.wide) <- make.names(names(dat_WFD_TW.wide)) # create valid names

#select only 2nd cycle assessments 2016
dat_WFD_TW_2nd <- dat_WFD_TW %>% 
  filter(cYear==2016) %>% 
  #filter(StartYear > 2009 |is.na(StartYear)) %>% 
  ungroup()

dat_WFD_TW_2nd_wide <- dat_WFD_TW_2nd %>% 
  select(euSurfaceWaterBodyCode,countryCode,qeCode,qeStatusOrPotentialValue,naturalAWBHMWB,surfaceWaterBodyTypeCode,
         surfaceWaterBodyIntercalibrationType,surfaceWaterBodyIntercalibrationTypeCode,cArea,cLength) %>% 
  pivot_wider(names_from = qeCode, values_from = qeStatusOrPotentialValue)

colnames(dat_WFD_TW_2nd_wide) <- make.names(names(dat_WFD_TW_2nd_wide)) # create valid names

#CW
dat_WFD_CW <- readRDS(here("Data","dat_W4_CW_BQE.rds")) #import biology

#check available reporting years
unique(dat_WFD_CW$cYear) # 2010, 2016

#all years
dat_WFD_CW.wide <- dat_WFD_CW %>% 
  select(euSurfaceWaterBodyCode,countryCode,qeCode,qeStatusOrPotentialValue,cYear,naturalAWBHMWB,surfaceWaterBodyTypeCode,
         surfaceWaterBodyIntercalibrationType,surfaceWaterBodyIntercalibrationTypeCode,cArea,cLength) %>% 
  pivot_wider(names_from = qeCode, values_from = qeStatusOrPotentialValue)

colnames(dat_WFD_CW.wide) <- make.names(names(dat_WFD_CW.wide)) # create valid names

#select only 2nd cycle assessments 2016
dat_WFD_CW_2nd <- dat_WFD_CW %>% 
  filter(cYear==2016) %>% 
  #filter(StartYear > 2009 |is.na(StartYear)) %>% 
  ungroup()

dat_WFD_CW_2nd_wide <- dat_WFD_CW_2nd %>% 
  select(euSurfaceWaterBodyCode,countryCode,qeCode,qeStatusOrPotentialValue,naturalAWBHMWB,surfaceWaterBodyTypeCode,
         surfaceWaterBodyIntercalibrationType,surfaceWaterBodyIntercalibrationTypeCode,cArea,cLength) %>% 
  pivot_wider(names_from = qeCode, values_from = qeStatusOrPotentialValue)

colnames(dat_WFD_CW_2nd_wide) <- make.names(names(dat_WFD_CW_2nd_wide)) # create valid names
```

```{r Import WISE2 Biota EQS classifications to Wide format}
#TW & CW SiteMean per year all period of data
dat_BQE_TW <- readRDS(here("Data","dat_BQE_TW.rds")) #import biology
dat_BQE_CW <- readRDS(here("Data","dat_BQE_CW.rds")) #import biology

#TRaC aggregated by WB per year all period of data
dat_BQE_TRaC.aggrWB <- readRDS(here("Data","dat_BQE_TRaC_aggrWB_all.rds")) #import biology aggregated by WB all years 

#TW
names(dat_BQE_TW)

dat_BQE_TW_wideEQS <- dat_BQE_TW %>% 
  select(countryCode,waterBodyIdentifier,waterBodyName,parameterNCSWaterBodyType,
         monitoringSiteIdentifier,monitoringSiteIdentifierScheme,parameterNaturalAWBHMWB,
         phenomenonTimeReferenceYear,observedPropertyDeterminandLabel,resultEcologicalStatusClassValue) %>% 
  pivot_wider(names_from = observedPropertyDeterminandLabel, 
              values_from = resultEcologicalStatusClassValue,
              values_fn = mean)   #how to combine duplicates

colnames(dat_BQE_TW_wideEQS) <- make.names(names(dat_BQE_TW_wideEQS)) # create valid names


#CW
dat_BQE_CW_wideEQS <- dat_BQE_CW %>% 
  select(countryCode,waterBodyIdentifier,waterBodyName,parameterNCSWaterBodyType,
         monitoringSiteIdentifier,monitoringSiteIdentifierScheme,parameterNaturalAWBHMWB,
         phenomenonTimeReferenceYear,observedPropertyDeterminandLabel,resultEcologicalStatusClassValue) %>% 
  pivot_wider(names_from = observedPropertyDeterminandLabel, 
              values_from = resultEcologicalStatusClassValue,
              values_fn = mean)

colnames(dat_BQE_CW_wideEQS) <- make.names(names(dat_BQE_CW_wideEQS)) # create valid names

#TRaC aggWB
dat_BQE_TRaC.aggrWB_wideEQS <- dat_BQE_TRaC.aggrWB %>% 
  select(countryCode,parameterWaterBodyCategory,waterBodyIdentifier,waterBodyName,parameterNCSWaterBodyType,
         waterBodyIdentifierScheme,parameterNaturalAWBHMWB,
         phenomenonTimeReferenceYear,observedPropertyDeterminandLabel,resultEcologicalStatusClassValue) %>% 
  pivot_wider(names_from = observedPropertyDeterminandLabel, values_from = resultEcologicalStatusClassValue)

colnames(dat_BQE_TRaC.aggrWB_wideEQS) <- make.names(names(dat_BQE_TRaC.aggrWB_wideEQS)) # create valid names

```

### B2. Combine WISE4 & WISE2 Biological EQS data for all available years
```{r harmonise BQEs names WISE2 & WISE4}
#wise4
dat_WFD_TW.wide #n=1119
dat_WFD_CW.wide #n=4735

#wise2
dat_BQE_TW_wideEQS #n=1496
dat_BQE_CW_wideEQS #n=3257
dat_BQE_TRaC.aggrWB_wideEQS #n=1451

#1st harmonise BQE names in both reporting databases 
names(dat_WFD_TW.wide)
names(dat_WFD_CW.wide)
names(dat_BQE_TW_wideEQS)
names(dat_BQE_CW_wideEQS)
names(dat_BQE_TRaC.aggrWB_wideEQS)

#TW 
dat_WFD_TW.wide <- dat_WFD_TW.wide %>%
  rename(
    InvertebrateEQR = QE1.3...Benthic.invertebrates,
    FishEQR = QE1.4...Fish,
    MacroalgaeEQR = QE1.2.1...Macroalgae,
    PhytoplanktonEQR = QE1.1...Phytoplankton,
    otherAquaticFlora = QE1.2...Other.aquatic.flora,
    Macrophytes = QE1.2.3...Macrophytes,
    AngiospermsEQR = QE1.2.2...Angiosperms,
    PhytobenthosEQR_E = QE1.2.4...Phytobenthos
  )

#CW
dat_WFD_CW.wide <- dat_WFD_CW.wide %>%
  rename(
    InvertebrateEQR = QE1.3...Benthic.invertebrates,
    FishEQR = QE1.4...Fish,
    MacroalgaeEQR = QE1.2.1...Macroalgae,
    PhytoplanktonEQR = QE1.1...Phytoplankton,
    otherAquaticFlora = QE1.2...Other.aquatic.flora,
    Macrophytes = QE1.2.3...Macrophytes,
    AngiospermsEQR = QE1.2.2...Angiosperms,
    PhytobenthosEQR_E = QE1.2.4...Phytobenthos
  )

```

```{r prepare data to merge BQE aggrWB from WISE2 & WISE4 }
#1st check for differences in variables names in wise 4 TW vs CW
setdiff(names(dat_WFD_TW.wide),names(dat_WFD_CW.wide)) #no difference between TW & CW WISE4

#2nd merge TW and CW in WISE4 into unique df
dat_WFD_TRaC.aggrWB_wideEQS <- bind_rows(TW = dat_WFD_TW.wide, CW = dat_WFD_CW.wide, 
                                         .id = "parameterWaterBodyCategory") #id source water category dataset

#3rd inspect WISE4 vs WISE2 TRaC aggregated by WB
names(dat_WFD_TRaC.aggrWB_wideEQS) #WISE4
names(dat_BQE_TRaC.aggrWB_wideEQS) #WISE2

intersect(names(dat_WFD_TRaC.aggrWB_wideEQS),names(dat_BQE_TRaC.aggrWB_wideEQS)) 
#equal variable names to match:
#"parameterWaterBodyCategory" "countryCode" 
#"InvertebrateEQR"  "FishEQR"    "MacroalgaeEQR"    "PhytoplanktonEQR" "AngiospermsEQR" 

symdiff(names(dat_WFD_TRaC.aggrWB_wideEQS),names(dat_BQE_TRaC.aggrWB_wideEQS))
#equivalent variables to match:
#"naturalAWBHMWB" =  "parameterNaturalAWBHMWB" 
#"euSurfaceWaterBodyCode" = "waterBodyIdentifier"
#"surfaceWaterBodyTypeCode" = "parameterNCSWaterBodyType"

#missing BQEs to include 
# "PhytobenthosEQR_E", "otherAquaticFlora", "Macrophytes"    

#diff vars ADD:
#"phenomenonTimeReferenceYear" ,  "cYear"  
#"waterBodyName", "surfaceWaterBodyIntercalibrationType","surfaceWaterBodyIntercalibrationTypeCode"

#diff vars DROP:
unique(dat_BQE_TRaC.aggrWB_wideEQS$waterBodyIdentifierScheme)     
# waterBodyIdentifierScheme : in wise 2 
# -cArea, -cLength : in wise 4

dat_WFD_TRaC.aggrWB_wideEQS <- dat_WFD_TRaC.aggrWB_wideEQS %>% #WISE4 clean
  select(-cArea, -cLength) 

# WISE2 clean data and aggregate by WFD Cycle BEFORE merge with WISE4
# add WFD Cycle year variable to wise 2 "cYear"
cYear <- list(
  `baseline`= 2000:2003,
  `2010` = 2004:2009,
  `2016` = 2010:2015,
  `2022` = 2016:2021,
  `2028` = 2022:2027
)

dat_BQE_TRaC.aggrWB_Cycle <- dat_BQE_TRaC.aggrWB_wideEQS %>% # wise2
  select(-waterBodyIdentifierScheme) %>%
  mutate(cYear=case_when(
    phenomenonTimeReferenceYear %in% cYear$baseline ~ 2000,
    phenomenonTimeReferenceYear %in% cYear$`2010` ~ 2010,
    phenomenonTimeReferenceYear %in% cYear$`2016` ~ 2016,
    phenomenonTimeReferenceYear %in% cYear$`2022` ~ 2022,
    phenomenonTimeReferenceYear %in% cYear$`2028` ~ 2028,
    TRUE ~ NA_real_
    )
    )

names(dat_BQE_TRaC.aggrWB_Cycle)

#4th summarise WISE2 BQE EQS data per per WB and WFD cycle
#to include mean EQS value rounded per BQE + count samples
meanEQS_TRaC_Cycle_WB_wise2 <- dat_BQE_TRaC.aggrWB_Cycle %>%
  group_by(parameterWaterBodyCategory, cYear, countryCode, parameterNCSWaterBodyType, waterBodyIdentifier, waterBodyName,parameterNaturalAWBHMWB) %>%
  summarise(
    # mean values for all EQS rounded to nearest integer, then clipped to 1–5 scale
    across(
      ends_with("EQR"),
      ~ {
        m <- mean(.x, na.rm = TRUE)
        if (is.nan(m)) return(NA_real_)     # keep NA
        m_round <- round(m, 0)
        m_safe  <- min(max(m_round, 1), 5) # force between 1 and 5
        m_safe
      },
      .names = "{.col}_mean"
    ),
    
    # count of non-missing records per EQS value
    across(
      ends_with("EQR"),
      ~ sum(!is.na(.x)),
      .names = "{.col}_n"
    ),
    
    .groups = "drop"
  )



#5th join tables WISE4 vs WISE2 TRaC aggregated by WB
names (meanEQS_TRaC_Cycle_WB_wise2) #WISE2
names(dat_WFD_TRaC.aggrWB_wideEQS) #WISE4

meanEQS_TRaC_Cycle_WB_wise2_clean<- meanEQS_TRaC_Cycle_WB_wise2 %>% 
  select(-PhytoplanktonEQR_n, -MacroalgaeEQR_n, -InvertebrateEQR_n, -AngiospermsEQR_n, -FishEQR_n) #drop count of samples to mean EQS in WISE2

symdiff(names(meanEQS_TRaC_Cycle_WB_wise2_clean),names(dat_WFD_TRaC.aggrWB_wideEQS)) #check all vars not shared between WISE4 & WISE2
 
  
```

```{r merge WISE4 & WISE2 TRaC aggregated by WB and per Cycle}
 
#merge WISE4 & WISE2 TRaC aggregated by WB full_join to keep all data
  dat_EQS_aggrWB_since2010  <- full_join(meanEQS_TRaC_Cycle_WB_wise2_clean, dat_WFD_TRaC.aggrWB_wideEQS,
          by=c("parameterWaterBodyCategory",
               "cYear",
               "countryCode",
               "waterBodyIdentifier"="euSurfaceWaterBodyCode", #establish equivalence where names differ
               "parameterNCSWaterBodyType"="surfaceWaterBodyTypeCode",
               "parameterNaturalAWBHMWB"="naturalAWBHMWB",
               "PhytoplanktonEQR_mean"="PhytoplanktonEQR",
               "MacroalgaeEQR_mean"="MacroalgaeEQR",
               "InvertebrateEQR_mean"="InvertebrateEQR",
               "AngiospermsEQR_mean" = "AngiospermsEQR",
               "FishEQR_mean"="FishEQR"
               )
  )
  
  #rename variables for accuracy, EQS not EQR
  dat_EQS_aggrWB_since2010 <- dat_EQS_aggrWB_since2010 %>%
    rename( 
    "PhytoplanktonEQS"="PhytoplanktonEQR_mean",
    "MacroalgaeEQS"="MacroalgaeEQR_mean",
    "InvertebrateEQS"="InvertebrateEQR_mean",
    "AngiospermsEQS"="AngiospermsEQR_mean",
    "FishEQS"="FishEQR_mean",
    "otherAquaticFloraEQS"="otherAquaticFlora",
    "MacrophytesEQS"= "Macrophytes",
    "PhytobenthosE_EQS"="PhytobenthosEQR_E"
  )
  
 #checks
 length(unique(dat_EQS_aggrWB_since2010$surfaceWaterBodyIntercalibrationTypeCode))
 unique(dat_EQS_aggrWB_since2010$surfaceWaterBodyIntercalibrationTypeCode)
 
 #RW observations misclassified as CW, dropped
    dat_EQS_aggrWB_since2010 <- dat_EQS_aggrWB_since2010 %>%
      filter(!surfaceWaterBodyIntercalibrationTypeCode %in% c("RW-R-M4", "RW-R-M1"))

#fill in missing ICtype code for  WISE2 periods using WISE4 available IC common Types codes per water body
    dat_EQS_aggrWB_since2010_ICtypes <- dat_EQS_aggrWB_since2010 %>%
  group_by(waterBodyIdentifier) %>%
  # Fill missing codes using available IC TypeCode for the same water body
  mutate(
    surfaceWaterBodyIntercalibrationTypeCode = 
      if_else(
        is.na(surfaceWaterBodyIntercalibrationTypeCode),
        first(na.omit(surfaceWaterBodyIntercalibrationTypeCode)),
        surfaceWaterBodyIntercalibrationTypeCode
      )
  ) %>%
  ungroup()
   
#check
dat_EQS_aggrWB_since2010_ICtypes %>% 
      group_by(waterBodyIdentifier,surfaceWaterBodyIntercalibrationTypeCode,cYear) %>%
    summarise(n=n(), .groups = "drop")%>%
      arrange(desc(cYear), desc(n))
```
